No need to know: astrophysics-agnostic gravitational wave cosmology

### 2024-03-22 src/tex/ms.tex
[showyourwork] overleaf sync

\DeclareAcronym{LVK}{short=LVK, long=LIGO Virgo KAGRA}
\DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}

@@

% Abstract
\begin{abstract}
\Acp{GW} from merging compact objects are the only transient sources that encode direct information about their luminosity distances. 
Paired with their redshifts, a Hubble diagram can be constructed to probe the Universe's expansion.  
``Spectral sirens'' use features in the mass distribution of \ac{GW} sources to extract this redshift information without the need for accompanying electromagnetic observations.
So far, this technique has only been applied with simple parametric representations of the mass distribution. 
However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.
Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters with the spectral siren methodology.
We demonstrate that prior knowledge of the \ac{GW} source mass distribution is not necessary in order to accurately infer cosmological parameters with spectral sirens.
We do this by constructing a flexible, Gaussian process-based model for the \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations and in the presence of redshift evolution of the mass spectrum.  
\end{abstract}

@@

Like light, \acp{GW} are redshifted as they propagate across the universe, thereby becoming imprinted with the cosmic expansion history.
Unlike light, however, the form of \ac{GW} signals are known from first principles and their selection effects are extremely well understood.
This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} population. 
Additionally, the \ac{GW} signals detected by the LIGO \citep{aasi_advanced_2015}, Virgo \citep{acernese_advanced_2014}, and KAGRA \citep{akutsu_overview_2021} detectors provide direct measurements of the distance to their sources.
This makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
This is particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.

A well-known application of standard sirens was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017}, whose clear association with a host galaxy provided a precise redshift measurement, allowing for a direct measurement of the Hubble constant, \Ho{} \citep{ abbott_gravitational-wave_2017}.
External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal, allowing for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.

However, external information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
In practice, unless tidal information is available \citep{messenger_measuring_2012}, the source frame mass is unknown.
However, it is possible to know the location of features in the source frame mass distribution.
The whole mass distribution therefore acts similarly to an electromagnetic spectrum, where the location of spectral features provides a redshift measurement.
The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993,Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}. 
Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest \ac{GW} catalog \cite{gwtc3_cosmo}.

Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff_gravitational_1993}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from theoretical expectations of e.g. the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass. 
This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
Such systematic bias is illustrated in Figure~\ref{fig:O5_GP}, where we show the result of fitting simulated data with a model that has the true mass distribution contained within its prior and one that does not. 
The latter does not accurately infer the true value of \Ho{}, whereas the former does.

@@

Because \ac{GW} observations are the only data input to spectral sirens, they are uniquely positioned to probe the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
We have shown this by simultaneously measuring \Omm{} and \Ho{}, but the method can be trivially extended to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$. 
We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ during \acf{O5}.
This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 
Proposed next generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (up to $z\sim100$).
At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.

>>>>

\DeclareAcronym{LVK}{short=LVK, long=LIGO-Virgo-KAGRA}
\DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}

@@

% Abstract
\begin{abstract}
\Acp{GW} from merging compact objects are the only transient sources that encode direct information about their luminosity distances. 
Once paired with their redshifts, a Hubble diagram can be constructed to directly probe the Universe's expansion.
This can be done in the absence of any electromagnetic information: features in the mass distribution of \ac{GW} can provide self-calibrating redshift measurements without the need for an accompanying electromagnetic transient or catalog of potential host galaxies.
\sout{So far, this technique has only been applied with simple parametric representations of the mass distribution. 
However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.}
It is commonly presumed, however, that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters in this fashion.
We demonstrate that prior knowledge of the \ac{GW} source mass distribution is not necessary in order to accurately infer cosmological parameters with spectral sirens.
By adopting a flexible, Gaussian-process model for the mass distribution of compact binaries, we find that we can accurately and agnostically learn this mass distribution while simultaneously measuring parameters governing cosmological expansion.
We demonstrate this analyses using a simulated catalog of $10^3$ gravitational-wave events consistent with expectations for the next \acl{LVK} observing run.
We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations and in the presence of redshift evolution of the mass spectrum.  
\end{abstract}

@@

Like light, \acp{GW} are redshifted as they propagate across the universe, thereby becoming imprinted with the cosmic expansion history.
Unlike light, however, the form of \ac{GW} signals are known from first principles and their selection effects are extremely well understood.
This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} source population~\citep{2023PhRvX..13d1039A,abbott_population_2023,2023PhRvD.108d3011E}. 
Additionally, the \ac{GW} signals observed by the LIGO, Virgo, and KAGRA detectors~\citep{aasi_advanced_2015,acernese_advanced_2014,akutsu_overview_2021}  provide direct measurements of the distance to their sources.
This makes them ``standard sirens'': direct probes of cosmological parameters that circumvent the need for a cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
This is particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.

Cosmological measurements with gravitational waves traditionally require independent electromagnetic measurement of source redshifts.
A well-known demonstration was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal and allow for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.

However, external electromagnetic information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.

In practice, the source-frame masses of individual events are not known~\citep[unless tidal information is available][]{messenger_measuring_2012}.
It is, however, possible to regard the \textit{population} of compact binaries at large and use known features in their source-frame mass distribution to obtain self-calibrated redshift estimates.
The full mass distribution therefore acts analogously to an electromagnetic spectrum, in which the apparent locations of spectral features provide a redshift measurement.
The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.

Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993} and \cite{Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}. 
Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest \ac{GW} catalog \citep{gwtc3_cosmo}.

Perhaps because of the analogy to electromagnetic spectra, in which the locations of source-frame spectral features are often known from first principles, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to perform a spectral siren redshift measurement.
Indeed, first principles models for the mass distribution of merging compact binaries are not available: significant theoretical uncertainties exist on the shapes, locations, and very existences of predicted features like a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap, or the maximum neutron star mass.
To this end, spectral siren cosmology\footnote{And also statistical cosmological measurements leveraging galaxy catalogs.} relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.

This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
Such systematic bias is illustrated in Figure~\ref{fig:O5_GP}, where we show the result of fitting simulated data with a model that has the true mass distribution contained within its prior and one that does not. 
The latter does not accurately infer the true value of \Ho{}, whereas the former does.

@@

Because \ac{GW} observations are the only data input to spectral sirens, they are uniquely positioned to probe the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
We have shown this by simultaneously measuring \Omm{} and \Ho{}, but the method can be trivially extended to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$. 
We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ during \acf{O5}.
This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
\comment{Projected DESI errors are $\sim3\%$ at $z=1$, which are comparable with our estimated error of $\variable{output/Hz_percent.txt}\%$, but those won't be realized until 202X.}
This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 
Proposed next generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (up to $z\sim100$).
At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
